---
title: "Network Traffic Analysis from AstraeaDB"
author: "Generated with R-AstraeaDB"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 7
)
```

# Overview

This document analyses IP traffic that was captured by `pcap_to_graph.R` and
stored in AstraeaDB.  The graph model is:

- **Nodes** (label `IPAddress`): one per unique IP, property `ip`.
- **Edges** (type = protocol): one per **aggregated flow**, defined by the
  tuple `(src_ip, dst_ip, protocol, service_port)`. Each edge carries:
    - `service_port` -- the classified server-side port
    - `service_name` -- human-readable name (e.g. "HTTPS"), if known
    - `flow_count`   -- number of packets aggregated into this flow
    - `total_bytes`  -- sum of packet sizes in the flow
    - `valid_from` / `valid_to` -- temporal window of the flow

```{r libraries}
library(AstraeaDB)
library(igraph)
library(ggplot2)
library(data.table)
library(scales)
library(visNetwork)

# Helper: flatten a list column to an atomic vector (NULL -> NA)
unbox_col <- function(x) {
  sapply(x, function(el) if (is.null(el) || length(el) == 0) NA else el[[1]])
}
```

# 1 -- Data Extraction

Connect to AstraeaDB, discover every IP node, and collect all flow edges.

```{r connect}
client <- astraea_connect()
client$ping()
```

```{r extract-nodes}
# Use GQL to find all IPAddress nodes regardless of their internal IDs.
node_result <- client$query("MATCH (n:IPAddress) RETURN n")
nodes_list <- lapply(node_result$rows, function(row) {
  n <- row[[1]]
  list(node_id = n$id, ip = n$properties$ip %||% NA_character_)
})

nodes_dt <- rbindlist(nodes_list, fill = TRUE)

# Flatten list columns to atomic types
for (col in names(nodes_dt)) {
  if (is.list(nodes_dt[[col]])) {
    set(nodes_dt, j = col, value = unbox_col(nodes_dt[[col]]))
  }
}
nodes_dt[, node_id := as.character(node_id)]
nodes_dt[, ip := as.character(ip)]

cat(sprintf("Discovered %d IP address nodes.\n", nrow(nodes_dt)))
```

```{r extract-edges}
# Use GQL to pull all flow edges between IPAddress nodes.
edge_result <- client$query(
  "MATCH (s:IPAddress)-[e]->(d:IPAddress)
   RETURN id(e), s.ip, d.ip, type(e), e.service_port, e.service_name,
          e.flow_count, e.total_bytes"
)

edges_list <- lapply(edge_result$rows, function(row) {
  list(
    edge_id      = row[[1]],
    src_ip       = row[[2]],
    dst_ip       = row[[3]],
    protocol     = row[[4]] %||% "IP",
    service_port = row[[5]] %||% NA_integer_,
    service_name = row[[6]] %||% NA_character_,
    flow_count   = row[[7]] %||% 1L,
    total_bytes  = row[[8]] %||% NA_integer_
  )
})

edges_dt <- rbindlist(edges_list, fill = TRUE)

# Flatten list columns to atomic types (query results may wrap values in lists)
for (col in names(edges_dt)) {
  if (is.list(edges_dt[[col]])) {
    set(edges_dt, j = col, value = unbox_col(edges_dt[[col]]))
  }
}
# Coerce to expected types
edges_dt[, edge_id      := as.character(edge_id)]
edges_dt[, src_ip       := as.character(src_ip)]
edges_dt[, dst_ip       := as.character(dst_ip)]
edges_dt[, protocol     := as.character(protocol)]
edges_dt[, service_port := as.integer(service_port)]
edges_dt[, service_name := as.character(service_name)]
edges_dt[, flow_count   := as.integer(flow_count)]
edges_dt[, total_bytes  := as.numeric(total_bytes)]

cat(sprintf("Collected %d flow edges (representing %s packets).\n",
            nrow(edges_dt),
            format(sum(edges_dt$flow_count, na.rm = TRUE), big.mark = ",")))
```

```{r extract-timestamps}
# Timestamps live in the edge's valid_from / valid_to fields.
edges_dt[, c("valid_from_ms", "valid_to_ms") := {
  vf <- vapply(edge_id, function(eid) {
    e <- tryCatch(client$get_edge(eid), error = function(e) NULL)
    if (!is.null(e) && !is.null(e$valid_from)) e$valid_from else NA_real_
  }, numeric(1))
  vt <- vapply(edge_id, function(eid) {
    e <- tryCatch(client$get_edge(eid), error = function(e) NULL)
    if (!is.null(e) && !is.null(e$valid_to)) e$valid_to else NA_real_
  }, numeric(1))
  list(vf, vt)
}]
```

```{r enrich}
# Build lookup and add derived columns.
ip_lookup <- setNames(nodes_dt$ip, as.character(nodes_dt$node_id))

edges_dt[, service_port_int := suppressWarnings(as.integer(service_port))]
edges_dt[, timestamp := as.POSIXct(valid_from_ms / 1000,
                                    origin = "1970-01-01")]
edges_dt[, timestamp_end := as.POSIXct(valid_to_ms / 1000,
                                        origin = "1970-01-01")]
edges_dt[, flow_duration_s := (valid_to_ms - valid_from_ms) / 1000]

# Use package function for service name lookup on missing names
edges_dt[is.na(service_name) & !is.na(service_port_int),
         service_name := port_service_name(service_port_int)]
```

```{r summary-table}
total_packets <- sum(edges_dt$flow_count, na.rm = TRUE)
summary_stats <- data.frame(
  Metric = c("Unique IP addresses",
             "Total flows (edges)",
             "Total packets (aggregated)",
             "Protocols observed",
             "Time span (seconds)",
             "Total bytes captured",
             "Avg packets per flow"),
  Value = c(
    nrow(nodes_dt),
    nrow(edges_dt),
    total_packets,
    length(unique(edges_dt$protocol)),
    if (nrow(edges_dt) > 0)
      round(as.numeric(difftime(max(edges_dt$timestamp, na.rm = TRUE),
                                min(edges_dt$timestamp, na.rm = TRUE),
                                units = "secs")), 1)
    else 0,
    sum(edges_dt$total_bytes, na.rm = TRUE),
    if (nrow(edges_dt) > 0) round(total_packets / nrow(edges_dt), 1) else 0
  )
)
knitr::kable(summary_stats, caption = "Capture summary")
```

# 2 -- Protocol Distribution

```{r protocol-bar}
proto_counts <- edges_dt[, .(
  flows   = .N,
  packets = sum(flow_count, na.rm = TRUE)
), by = protocol][order(-packets)]

ggplot(proto_counts, aes(x = reorder(protocol, packets),
                         y = packets, fill = protocol)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(title = "Packets by Protocol (aggregated from flows)",
       x = "Protocol", y = "Packet count") +
  theme_minimal(base_size = 14)
```

```{r protocol-bytes}
proto_bytes <- edges_dt[, .(
  total_bytes  = sum(total_bytes, na.rm = TRUE),
  total_pkts   = sum(flow_count, na.rm = TRUE),
  avg_pkt_size = round(sum(total_bytes, na.rm = TRUE) /
                         sum(flow_count, na.rm = TRUE)),
  flow_count   = .N
), by = protocol][order(-total_bytes)]

knitr::kable(proto_bytes, caption = "Traffic volume by protocol",
             format.args = list(big.mark = ","))
```

# 3 -- Top Talkers

## By total packets sent

```{r top-senders}
top_senders <- edges_dt[, .(packets_sent = sum(flow_count, na.rm = TRUE)),
                        by = src_ip][order(-packets_sent)][1:min(.N, 20)]

ggplot(top_senders, aes(x = reorder(src_ip, packets_sent), y = packets_sent)) +
  geom_col(fill = "#3498db") +
  coord_flip() +
  labs(title = "Top 20 Source IPs by Packet Count",
       x = NULL, y = "Packets sent") +
  theme_minimal(base_size = 13)
```

## By total bytes sent

```{r top-senders-bytes}
top_bytes <- edges_dt[, .(total_bytes = sum(total_bytes, na.rm = TRUE)),
                      by = src_ip][order(-total_bytes)][1:min(.N, 20)]

ggplot(top_bytes, aes(x = reorder(src_ip, total_bytes), y = total_bytes)) +
  geom_col(fill = "#2ecc71") +
  coord_flip() +
  labs(title = "Top 20 Source IPs by Bytes Sent",
       x = NULL, y = "Bytes") +
  scale_y_continuous(labels = label_bytes()) +
  theme_minimal(base_size = 13)
```

## Busiest conversations (IP pairs)

```{r top-pairs}
pair_traffic <- edges_dt[, .(
  flows   = .N,
  packets = sum(flow_count, na.rm = TRUE),
  bytes   = sum(total_bytes, na.rm = TRUE)
), by = .(src_ip, dst_ip)][order(-packets)]

knitr::kable(head(pair_traffic, 20),
             caption = "Top 20 IP-pair conversations by packet count",
             format.args = list(big.mark = ","))
```

# 4 -- Port Analysis

## Service Port Distribution

```{r service-port-dist}
svc_counts <- edges_dt[!is.na(service_port_int), .(
  flows   = .N,
  packets = sum(flow_count, na.rm = TRUE),
  bytes   = sum(total_bytes, na.rm = TRUE)
), by = .(service_port_int, service_name)][order(-packets)]

svc_counts[, label := fifelse(
  !is.na(service_name) & service_name != "",
  paste0(service_port_int, " (", service_name, ")"),
  as.character(service_port_int)
)]

top_svcs <- svc_counts[1:min(.N, 20)]

ggplot(top_svcs, aes(x = reorder(label, packets), y = packets)) +
  geom_col(fill = "#9b59b6") +
  coord_flip() +
  labs(title = "Top 20 Service Ports by Packet Count",
       x = NULL, y = "Packet count (aggregated)") +
  theme_minimal(base_size = 13)
```

```{r service-port-bytes}
ggplot(top_svcs, aes(x = reorder(label, bytes), y = bytes)) +
  geom_col(fill = "#e67e22") +
  coord_flip() +
  labs(title = "Top 20 Service Ports by Total Bytes",
       x = NULL, y = "Total bytes") +
  scale_y_continuous(labels = label_bytes()) +
  theme_minimal(base_size = 13)
```

## Port-protocol cross-tabulation

```{r port-proto-heatmap, fig.height=8}
pp <- edges_dt[!is.na(service_port_int), .(
  packets = sum(flow_count, na.rm = TRUE)
), by = .(protocol, service_port_int)][order(-packets)]
pp_top <- pp[service_port_int %in% svc_counts$service_port_int[1:15]]

if (nrow(pp_top) > 0) {
  ggplot(pp_top, aes(x = protocol,
                     y = factor(service_port_int),
                     fill = log10(packets + 1))) +
    geom_tile(color = "white") +
    scale_fill_viridis_c(name = "log10(packets)") +
    labs(title = "Protocol vs Service Port",
         x = "Protocol", y = "Service port") +
    theme_minimal(base_size = 13)
}
```

# 5 -- Temporal Patterns

```{r time-series, fig.height=5}
if (sum(!is.na(edges_dt$timestamp)) > 0) {
  ts_dt <- edges_dt[!is.na(timestamp)]
  # Bin flow arrivals into 1-second intervals, weighting by flow_count
  ts_dt[, second := as.POSIXct(floor(as.numeric(timestamp)),
                                origin = "1970-01-01")]
  ts_agg <- ts_dt[, .(packets = sum(flow_count, na.rm = TRUE),
                       bytes = sum(total_bytes, na.rm = TRUE),
                       flows = .N),
                  by = second][order(second)]

  ggplot(ts_agg, aes(x = second, y = packets)) +
    geom_line(color = "#2c3e50", linewidth = 0.5) +
    geom_smooth(method = "loess", se = TRUE, color = "#e74c3c", span = 0.3) +
    labs(title = "Packet Rate Over Time (from flow aggregates)",
         x = "Time", y = "Packets per second") +
    theme_minimal(base_size = 13)
}
```

```{r flow-arrival-rate, fig.height=5}
if (exists("ts_agg") && nrow(ts_agg) > 0) {
  ggplot(ts_agg, aes(x = second, y = flows)) +
    geom_line(color = "#1abc9c", linewidth = 0.5) +
    geom_smooth(method = "loess", se = TRUE, color = "#8e44ad", span = 0.3) +
    labs(title = "Flow Arrival Rate Over Time",
         x = "Time", y = "New flows per second") +
    theme_minimal(base_size = 13)
}
```

```{r time-by-proto, fig.height=6}
if (sum(!is.na(edges_dt$timestamp)) > 0) {
  ts_proto <- ts_dt[, .(packets = sum(flow_count, na.rm = TRUE)),
                    by = .(second, protocol)][order(second)]

  ggplot(ts_proto, aes(x = second, y = packets, fill = protocol)) +
    geom_area(alpha = 0.7, position = "stack") +
    labs(title = "Traffic Volume by Protocol Over Time",
         x = "Time", y = "Packets per second", fill = "Protocol") +
    theme_minimal(base_size = 13)
}
```

```{r flow-duration, fig.height=5}
if (sum(!is.na(edges_dt$flow_duration_s)) > 0) {
  ggplot(edges_dt[!is.na(flow_duration_s) & flow_duration_s > 0],
         aes(x = flow_duration_s)) +
    geom_histogram(fill = "#2c3e50", color = "white", bins = 50) +
    scale_x_log10() +
    labs(title = "Flow Duration Distribution",
         x = "Duration (seconds, log scale)", y = "Number of flows") +
    theme_minimal(base_size = 13)
}
```

# 6 -- Network Graph Analysis

Build an `igraph` network where edge weights are packet counts between IP pairs.

```{r build-igraph}
# Aggregate to one edge per (src, dst, protocol) tuple, summing flow counts
agg <- edges_dt[, .(
  weight      = sum(flow_count, na.rm = TRUE),
  total_bytes = sum(total_bytes, na.rm = TRUE),
  num_flows   = .N
), by = .(src_ip, dst_ip, protocol)]

# Use only IPs that appear in edges (some nodes may be isolates)
all_ips <- unique(c(edges_dt$src_ip, edges_dt$dst_ip, nodes_dt$ip))
g <- graph_from_data_frame(
  d        = agg[, .(from = src_ip, to = dst_ip,
                      weight = weight, protocol = protocol,
                      total_bytes = total_bytes,
                      num_flows = num_flows)],
  directed = TRUE,
  vertices = data.frame(name = all_ips, stringsAsFactors = FALSE)
)

cat(sprintf("Graph: %d vertices, %d aggregated edges\n",
            vcount(g), ecount(g)))
```

## Degree distribution

```{r degree-dist}
deg_in  <- degree(g, mode = "in")
deg_out <- degree(g, mode = "out")
deg_all <- degree(g, mode = "all")

deg_df <- data.frame(
  ip     = names(deg_all),
  degree_in  = deg_in,
  degree_out = deg_out,
  degree_total = deg_all,
  stringsAsFactors = FALSE
)

ggplot(data.frame(degree = deg_all), aes(x = degree)) +
  geom_histogram(fill = "#1abc9c", color = "white", bins = 30) +
  labs(title = "Degree Distribution (all directions)",
       x = "Degree", y = "Number of IPs") +
  theme_minimal(base_size = 13)
```

```{r degree-log, fig.height=5}
# Check for power-law / heavy tail
deg_tab <- as.data.frame(table(degree = deg_all))
deg_tab$degree <- as.integer(as.character(deg_tab$degree))
deg_tab <- deg_tab[deg_tab$degree > 0, ]

if (nrow(deg_tab) > 3) {
  ggplot(deg_tab, aes(x = degree, y = Freq)) +
    geom_point(color = "#e67e22", size = 2) +
    scale_x_log10() + scale_y_log10() +
    labs(title = "Degree Distribution (log-log scale)",
         subtitle = "A straight line suggests scale-free / power-law structure",
         x = "Degree (log)", y = "Count (log)") +
    theme_minimal(base_size = 13)
}
```

## Centrality measures

```{r centrality}
# Compute centralities on the simplified (no multi-edge) graph
gs <- simplify(g, edge.attr.comb = list(weight = "sum",
                                         total_bytes = "sum",
                                         num_flows = "sum",
                                         protocol = "first"))

centrality_dt <- data.table(
  ip        = V(gs)$name,
  degree    = degree(gs, mode = "all"),
  in_degree = degree(gs, mode = "in"),
  out_degree = degree(gs, mode = "out"),
  strength  = strength(gs, mode = "all"),
  betweenness = round(betweenness(gs, directed = TRUE, normalized = TRUE), 6),
  pagerank  = round(page_rank(gs, directed = TRUE)$vector, 6)
)
setorder(centrality_dt, -pagerank)

knitr::kable(head(centrality_dt, 20),
             caption = "Top 20 IPs by PageRank")
```

## Community detection

```{r communities, fig.height=8}
g_undir <- as.undirected(gs, mode = "collapse",
                         edge.attr.comb = list(weight = "sum",
                                               total_bytes = "sum",
                                               num_flows = "sum",
                                               protocol = "first"))
if (vcount(g_undir) >= 3) {
  comm <- cluster_louvain(g_undir, weights = E(g_undir)$weight)
  cat(sprintf("Louvain detected %d communities (modularity = %.3f)\n",
              length(comm), modularity(comm)))

  comm_dt <- data.table(
    ip = V(g_undir)$name,
    community = membership(comm)
  )
  comm_sizes <- comm_dt[, .N, by = community][order(-N)]
  setnames(comm_sizes, "N", "members")

  knitr::kable(comm_sizes, caption = "Community sizes")

  # Store membership for visualisation
  V(g_undir)$community <- membership(comm)
}
```

# 7 -- Anomaly Indicators

Flag IPs that exhibit unusual behaviour: high fan-out to many destinations
(potential scanning), connections to many distinct service ports (potential
reconnaissance), or asymmetric traffic ratios.

```{r anomalies}
anomaly_dt <- edges_dt[, .(
  unique_dst_ips   = uniqueN(dst_ip),
  unique_svc_ports = uniqueN(service_port_int, na.rm = TRUE),
  packets_sent     = sum(flow_count, na.rm = TRUE),
  bytes_sent       = sum(total_bytes, na.rm = TRUE),
  flows_out        = .N
), by = src_ip]

# Incoming stats
incoming <- edges_dt[, .(
  packets_received = sum(flow_count, na.rm = TRUE),
  bytes_received   = sum(total_bytes, na.rm = TRUE),
  flows_in         = .N
), by = .(dst_ip)]

anomaly_dt <- merge(anomaly_dt, incoming,
                    by.x = "src_ip", by.y = "dst_ip", all.x = TRUE)
anomaly_dt[is.na(packets_received), packets_received := 0L]
anomaly_dt[is.na(bytes_received), bytes_received := 0L]
anomaly_dt[is.na(flows_in), flows_in := 0L]

# Asymmetry ratio: > 1 means more sending than receiving
anomaly_dt[, send_recv_ratio := fifelse(
  packets_received > 0,
  round(packets_sent / packets_received, 2),
  Inf
)]
```

## Potential scanners (high destination fan-out)

```{r scanners}
scanner_thresh <- quantile(anomaly_dt$unique_dst_ips, 0.9, na.rm = TRUE)
scanners <- anomaly_dt[unique_dst_ips >= max(scanner_thresh, 3)][order(-unique_dst_ips)]

if (nrow(scanners) > 0) {
  ggplot(scanners, aes(x = reorder(src_ip, unique_dst_ips),
                       y = unique_dst_ips)) +
    geom_col(fill = "#e74c3c") +
    coord_flip() +
    labs(title = "IPs with High Destination Fan-Out",
         subtitle = "Potential network scanning behaviour",
         x = NULL, y = "Unique destination IPs") +
    theme_minimal(base_size = 13)
}

knitr::kable(head(scanners, 15),
             caption = "IPs contacting the most unique destinations",
             format.args = list(big.mark = ","))
```

## Port reconnaissance (many distinct service ports)

```{r port-recon}
recon_thresh <- quantile(anomaly_dt$unique_svc_ports, 0.9, na.rm = TRUE)
recon <- anomaly_dt[unique_svc_ports >= max(recon_thresh, 5)][order(-unique_svc_ports)]

if (nrow(recon) > 0) {
  ggplot(recon, aes(x = reorder(src_ip, unique_svc_ports),
                    y = unique_svc_ports)) +
    geom_col(fill = "#f39c12") +
    coord_flip() +
    labs(title = "IPs Targeting Many Distinct Service Ports",
         subtitle = "Potential port reconnaissance",
         x = NULL, y = "Unique service ports") +
    theme_minimal(base_size = 13)
}
```

## Send/receive asymmetry

```{r asymmetry}
asym <- anomaly_dt[is.finite(send_recv_ratio)][order(-send_recv_ratio)]

if (nrow(asym) > 0) {
  ggplot(asym, aes(x = packets_sent, y = packets_received,
                   color = send_recv_ratio)) +
    geom_point(size = 3, alpha = 0.8) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                color = "grey50") +
    scale_color_viridis_c(name = "Send/Recv\nratio", trans = "log1p") +
    labs(title = "Traffic Symmetry per IP",
         subtitle = "Points above the dashed line receive more than they send",
         x = "Packets sent", y = "Packets received") +
    theme_minimal(base_size = 13)
}
```

# 8 -- Interactive Network Map

```{r visnetwork, fig.height=8}
# Prepare visNetwork data from the simplified igraph
vis_nodes <- data.frame(
  id    = V(gs)$name,
  label = V(gs)$name,
  stringsAsFactors = FALSE
)

# Size nodes by total strength (packets in + out)
node_strength <- strength(gs, mode = "all")
vis_nodes$value <- as.numeric(node_strength[vis_nodes$id])

# Colour by community if available
if (!is.null(V(g_undir)$community)) {
  comm_map <- setNames(V(g_undir)$community, V(g_undir)$name)
  vis_nodes$group <- as.character(comm_map[vis_nodes$id])
}

vis_edges <- data.frame(
  from   = ends(gs, E(gs))[, 1],
  to     = ends(gs, E(gs))[, 2],
  value  = E(gs)$weight,
  title  = paste0(E(gs)$protocol, " — ",
                  format(E(gs)$weight, big.mark = ","), " packets, ",
                  format(E(gs)$total_bytes, big.mark = ","), " bytes"),
  arrows = "to",
  stringsAsFactors = FALSE
)

visNetwork(vis_nodes, vis_edges,
           main = "Captured IP Traffic Graph",
           submain = paste(nrow(vis_nodes), "hosts,",
                           nrow(vis_edges), "aggregated connections")) %>%
  visOptions(highlightNearest = list(enabled = TRUE, degree = 1),
             nodesIdSelection = TRUE) %>%
  visPhysics(solver = "forceAtlas2Based",
             forceAtlas2Based = list(gravitationalConstant = -50)) %>%
  visInteraction(navigationButtons = TRUE)
```

# 9 -- Payload Analysis

Examine the distribution of traffic volumes across flows to identify traffic
profiles (e.g., small DNS lookups vs large data transfers).

```{r avg-pkt-size}
edges_dt[, avg_pkt_size := fifelse(
  flow_count > 0,
  total_bytes / flow_count,
  NA_real_
)]
```

```{r payload-dist}
ggplot(edges_dt[!is.na(total_bytes)],
       aes(x = total_bytes)) +
  geom_histogram(fill = "#2c3e50", color = "white", bins = 50) +
  scale_x_log10(labels = label_bytes()) +
  labs(title = "Flow Size Distribution",
       x = "Total bytes per flow (log scale)", y = "Number of flows") +
  theme_minimal(base_size = 13)
```

```{r avg-pkt-by-proto, fig.height=6}
if (edges_dt[, uniqueN(protocol)] > 1) {
  ggplot(edges_dt[!is.na(avg_pkt_size)],
         aes(x = protocol, y = avg_pkt_size, fill = protocol)) +
    geom_violin(show.legend = FALSE, alpha = 0.7) +
    geom_boxplot(width = 0.15, outlier.size = 0.5, show.legend = FALSE) +
    labs(title = "Average Packet Size by Protocol",
         subtitle = "total_bytes / flow_count for each flow",
         x = "Protocol", y = "Average packet size (bytes)") +
    theme_minimal(base_size = 13)
}
```

```{r flow-count-dist}
ggplot(edges_dt[flow_count > 0],
       aes(x = flow_count)) +
  geom_histogram(fill = "#3498db", color = "white", bins = 50) +
  scale_x_log10() +
  labs(title = "Packets per Flow Distribution",
       x = "Packets per flow (log scale)", y = "Number of flows") +
  theme_minimal(base_size = 13)
```

# 10 -- AstraeaDB Graph Traversal Examples

Demonstrate using AstraeaDB's built-in graph algorithms on the captured data.

```{r bfs-example}
# BFS from the highest-PageRank node
top_ip     <- centrality_dt$ip[1]
top_ip_id  <- nodes_dt[ip == top_ip, node_id]
if (length(top_ip_id) == 1) {
  bfs_result <- client$bfs(top_ip_id, max_depth = 2L)
  cat(sprintf(
    "BFS from %s (node %d): reached %d nodes within 2 hops.\n",
    top_ip, top_ip_id, length(bfs_result)
  ))
}
```

```{r shortest-path-example}
# Shortest path between the two highest-PageRank nodes
if (nrow(centrality_dt) >= 2) {
  id_a <- nodes_dt[ip == centrality_dt$ip[1], node_id]
  id_b <- nodes_dt[ip == centrality_dt$ip[2], node_id]
  if (length(id_a) == 1 && length(id_b) == 1) {
    sp <- tryCatch(
      client$shortest_path(id_a, id_b, weighted = FALSE),
      error = function(e) NULL
    )
    if (!is.null(sp) && length(sp$path) > 0) {
      path_ips <- sapply(sp$path, function(nid) ip_lookup[as.character(nid)])
      cat(sprintf("Shortest path (%d hops): %s\n",
                  length(sp$path) - 1L,
                  paste(path_ips, collapse = " -> ")))
    } else {
      cat("No path found between the top two IPs.\n")
    }
  }
}
```

---

```{r disconnect, include=FALSE}
client$disconnect()
```

*Analysis generated on `r Sys.time()` from AstraeaDB captured traffic.*
